SFJ: An implementation of Semantic Featherweight Java

### 2020-04-21 conference/camera ready/paper.tex
second algorithm entered

\documentclass[runningheads]{llncs}
%%%%%%%%%%%%%%%%%%%%%%%%
%PACKAGES
%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}

@@

% Algorithm typesetting package
\usepackage[plain, shortend, linesnumberedhidden, ruled]{algorithm2e}
\SetAlFnt{\sffamily \small}
\SetKwRepeat{Do}{do}{while}
\SetKwProg{Fn}{Function}{:}{end}
\SetKwInOut{Input}{input}\SetKwInOut{Output}{output}

%% For section renaming in autoref
\addto\extrasenglish{%
  \renewcommand{\sectionautorefname}{\S}%
  \renewcommand{\subsectionautorefname}{\S}%
  \renewcommand{\algorithmautorefname}{Algo.}%
}

@@

\section{SFJ: Design and Implementation}
\label{sec:design}
Since we want to use types $\tau$ in practice in SFJ, we restrict them to \emph{finite trees} whose leaves are basic types, $\btypes$ \autoref{sec:terms} with no cycles.
For example, a recursive type $\alpha = [a : \alpha]$ denotes an infinite program tree \textbf{new} $C$(\textbf{new} $C$($\cdots{}$)), hence we rule it out as it is not inhabitable. Similarly, the types $\alpha = [b: \beta]$, $\beta = [a = \alpha]$ create a cycle and thus would not be inhabitable.
Notice that the above types can be defined in Java but as well are not inhabitable.

The type system of SFJ builds upon the type system by Dardha \emph{et al.} \cite{Dardha2013,Dardha2017}.
It is implemented using ANTLR \cite{parr2013}. We start by defining the grammar of the language in Extended Backus-Naur Form (EBNF), following \autoref{sec:types} and by using ANTLR, we can automatically generate a parser for SFJ and extend it in order to implement the required checks for our type system. Running the parser returns an abstract syntax tree (AST) of a program in SFJ.

When visiting the AST of a program, we check if it is well-formed, following the intuition at the end of \autoref{sec:background}. We mark any classes containing fields typed with only basic types as \emph{resolved} otherwise, as \emph{unresolved}.
Using this information, \autoref{alg:types} checks if the type definitions in a program are valid, namely, if they are finite trees whose leaves are basic types with no cycles.
%
At each iteration of the algorithm we resolve at least one type, until all the types are resolved, and return $True$. However, if there is at least one type unresolved, that means there is a cycle in the type definition and we return $False$.

\begin{algorithm}[t]
    \SetAlgoLined
    \caption{ValidityCheck}
    \label{alg:types}

    \DontPrintSemicolon
    \Input{classes marked $resolved/unresolved$ depending if their fields contain only basic types.}
    \Output{$True$ if all classes are valid type definitions, $False$ otherwise.}
    \BlankLine
    \Begin{
        \Do{$resolutionOccured = true$}{ %\tcc {comment}
            $resolutionOccured \longleftarrow false$ \;
            \For{class that is unresolved in classes}{
                $resolved \longleftarrow true$ \;
                \For{field in class that contains a class type}{
                    \If{type of field is unresolved}{
                        $resolved \longleftarrow false$
                    }

@@

%
Now, we can define the subtyping relation.
Building upon the interpretation of types as sets of values to define the semantic subtyping for FJ \cite{Dardha2013,Dardha2017}, in our implementation of SFJ we keep track of the subtyping relation by defining a map from a type to the set of its subtypes, with the property that the set of values of a subtype is included in the set of values of the type.
As a first step, we start with basic types.
\setlength{\abovedisplayskip}{5pt}
\setlength{\belowdisplayskip}{5pt}
\begin{flalign*}
    &Double  = \left\{Double, Float, Int, Short, Byte\right\} && Float = \left\{Float, Short, Byte\right\}\\
    &Long = \left\{Long, Int, Short, Byte\right\}  && Int = \left\{Int, Short, Byte\right\}\\
    &Short = \left\{Short, Byte\right\} \qquad Byte = \left\{Byte\right\} && Boolean = \left\{Boolean \right\}
\end{flalign*}

Note that \emph{Int} is not a subtype of \emph{Float} as a 32-bit \emph{float} cannot represent the whole set of 32-bit \emph{integer} values accurately and therefore \emph{Int} is not fully set-contained, however this is not the case for \emph{Int} and \emph{Double}.
Similarly, \emph{Long} is not a subtype of \emph{Double}.

Finally, the following algorithm defines the subtyping relation for all class types, which concludes the process.
\begin{verbatim}
function generateRelation(classes):
    List<class> untyped = []
    for class in classes:
        if addClass(class) is false:
            untyped.add(class)
    if untyped is not []:
        generateRelation(untyped)

function addClass(class):
    for existing class type in relation:
        if checkSuperSet(class, existingClass) is false:
            return false
        checkSuperSet(existingClass, class)
        add class to its own subtype relation

function checkSuperSet(class, other):
    boolean flag = true
    for field in class:
        if field contains type not in relation:
            return false
        if other does not have field:
            flag = false
        else:
            if other.field.types not fully contains field.types:
                flag = false
    for method in class:
        if method contains type not in relation:
            return false
        if other does not have method:
            flag = false
        else:
            if other.method.types not fully contains method.types:
                flag = false
    if flag == true:
        add class to other subtype relation
\end{verbatim}
Let us illustrate the subtyping algorithm with our \emph{Polygons} \autoref{polygons}.
The algorithm generates the following subtyping relation, which also includes the subtyping relation for basic types defined in the first step.
\begin{flalign*}
    &Polygon = \left\{Polygon, Triangle, Square, Rhombus \right\} \qquad Triangle = \left\{Triangle\right\}\\
    &Square = \left\{Square\right\} \qquad  Rhombus  = \left\{Rhombus\right\} \qquad
    Diagonal = \left\{Diagonal\right\}
\end{flalign*}
Recall method \emph{diagonal} in class \emph{Diagonal}, we can see that the result of the set operation on its parameter type gives the following set of polygons:
\begin{flalign*}
    Polygon \textbf{ and not } Triangle = \left\{Polygon, Square, Rhombus \right\}
\end{flalign*}
If we pass to \emph{diagonal} an argument of type \emph{Triangle}, as it is not contained in the computed set of subtypes, we get a typechecking error at compile time \cite{UD20}.

>>>>

\documentclass[runningheads]{llncs}
%%%%%%%%%%%%%%%%%%%%%%%%
%PACKAGES
%%%%%%%%%%%%%%%%%%%%%%%%
%\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}

@@

\renewcommand{\theequation}{\arabic{section}.\arabic{equation}}

% Algorithm typesetting package
\usepackage[plain, shortend, linesnumberedhidden, ruled]{algorithm2e}
\SetAlFnt{\sffamily \small}
\SetKwRepeat{Do}{do}{while}
\SetKwProg{Fn}{Function}{:}{end}
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}

%% For section renaming in autoref
\addto\extrasenglish{%
  \renewcommand{\sectionautorefname}{\S}%
  \renewcommand{\subsectionautorefname}{\S}%
  \renewcommand{\algorithmautorefname}{Algorithm}%
  %\renewcommand{\exampleautorefname}{Ex.}%
}

@@

\section{SFJ: Design and Implementation}
\label{sec:design}
\subsection{Valid type definitions}
Since we want to use types $\tau$ in practice in SFJ, we restrict them to \emph{finite trees} whose leaves are basic types $\btypes$ \autoref{sec:terms} with no cycles.
For example, a recursive type $\alpha = [a : \alpha]$ denotes an infinite program tree \textbf{new} $C$(\textbf{new} $C$($\cdots{}$)), hence we rule it out as it is not inhabitable. Similarly, the types $\alpha = [b: \beta]$, $\beta = [a = \alpha]$ create a cycle and thus would not be inhabitable.
Notice that the above types can be defined in Java, but as well are not inhabitable.

SFJ is implemented using ANTLR \cite{parr2013}. We start by defining the grammar of the language in Extended Backus-Naur Form (EBNF), following \autoref{sec:types} and by using ANTLR, we can automatically generate a parser for SFJ and extend it in order to implement the required checks for our types and type system. Running the parser on an SFJ program returns an abstract syntax tree (AST) of that program. 
%
When visiting the AST, we check if the program is well-formed, following the intuition at the end of \autoref{sec:background}. We mark any classes containing fields typed with only basic types as \emph{resolved} otherwise, as \emph{unresolved}.
Using this information, \autoref{alg:types} checks if the type definitions in a program are valid, namely, if they are finite trees whose leaves are basic types with no cycles.
%
At each iteration of the algorithm we can resolve at least one type, until all the types are resolved, and return $True$. However, if there is at least one type unresolved, that means there is a cycle in the type definition, and we return $False$.

\begin{algorithm}[t]
    \SetAlgoLined
    \caption{Validity Check for Type Definitions}
    \label{alg:types}

    \DontPrintSemicolon
    \Input{$classes$, the set of classes marked $resolved/unresolved$ depending if their fields contain only basic types.}
    \Output{$True$ if all classes are valid type definitions, $False$ otherwise.}
    \BlankLine
    \Begin{
        \Do{$resolutionOccured = true$}{ %\tcc {comment}
            $resolutionOccured \longleftarrow false$ \;
            \For{class that is unresolved in classes}{
                $resolved \longleftarrow true$ \;
                \For{field in class that contains a class type}{
                    \If{type of field is unresolved}{
                        $resolved \longleftarrow false$
                    }

@@

\subsection{Building semantic subtyping for SFJ}
If \autoref{alg:types} returns $True$, we can proceed to building semantic subtyping for SFJ.
Leveraging the interpretation of types as sets of values to define the semantic subtyping for FJ \cite{Dardha2013,Dardha2017}, in our implementation of SFJ we keep track of the semantic subtyping relation by defining a map from a type to the set of its subtypes, satisfying the property that the set of values of a subtype is included in the set of values of the type.
We start with basic types and define $Universe$ to be the set of all types.
The mapping for basic types is defined as follows.
\begin{equation}
    \label{eq:initialSubtype}
    \begin{array}{llllll}
        & Double   & = \left\{Double, Float, Int, Short, Byte\right\} &  & Float & = \left\{Float, Short, Byte\right\} \\
        & Long     & =    \left\{Long, Int, Short, Byte\right\}       &  & Int   & = \left\{Int, Short, Byte\right\}   \\
        & Short    & =    \left\{Short, Byte\right\}                  &  & Byte  & = \left\{Byte\right\}               \\
        & Boolean  & =    \left\{Boolean\right\}                      &  & Void  & = \left\{Void\right\}               \\
        & Universe & = \{Double, Float, Long, Int,                                                                     \\
        &          & \quad Short, Byte, Boolean, Void\}
   \end{array}
\end{equation}

Note that \emph{Int} is not a subtype of \emph{Float} as a 32-bit \emph{float} cannot represent the whole set of 32-bit \emph{integer} values accurately and therefore \emph{Int} is not fully set-contained in \emph{Float}, however this is not the case for \emph{Int} and \emph{Double}.
Similarly, \emph{Long} is not a subtype of \emph{Double}.

\autoref{alg:subtyping} builds the semantic subtyping relation for all class types of an SFJ program by calling the function \textit{generateRelation}. Given that classes are valid type definitions by \autoref{alg:types}, we are guaranteed that \autoref{alg:subtyping} will terminate.

%
\begin{algorithm}
    \caption{Semantic Subtyping for SFJ}
    \label{alg:subtyping}

    \DontPrintSemicolon
    \Input{$classes$, the set of classes in an SFJ program for which we have not yet defined a subtyping relation.\\
    $relation$, mapping of types to the set of subtypes. Initially this relation\\ only contains the entries defined in Equation \ref{eq:initialSubtype}. \;
    }
    \Begin{
        \Fn{generateRelation(classes: List<Class>)}{
            $unprocessed: List<Class>\ \longleftarrow []$

            \For{class in classes}{
                \If{addClass(class) = $False$}{
                    $unprocessed.add(class)$
                }
            }
            \If{$untyped \neq []$}{
                $generateRelation(unprocessed)$
            }
        }
        \;
        \Fn{addClass(class: Class) $\rightarrow$ boolean}{
            \For{existing class type in relation}{
                \If{checkSuperSet(class, existingClass) = false}{
                    \KwRet $False$\;
                }
                $checkSuperSet(existingClass, class)$\;
            }
            $relation[class].add(class)$ \;
            $relation[Universe].add(class)$ \;
            \KwRet $True$ \;
        }
        \;
        \Fn{checkSuperSet(class: Class, other: Class) $\rightarrow$ boolean}{
            $flag \longleftarrow True$ \;
            \For{field in class}{
                \If{field contains type not in relation}{
                    \KwRet $False$
                }
                \uIf{other does not contain field}{
                    $flag \longleftarrow False$
                }
                \Else{
                    \If{other.field.types does not fully contain field.types}{
                        $flag \longleftarrow False$
                    }
                }
            }

            \For{method in class}{
                \If{method contains type not in relation}{
                    \KwRet $False$
                }
                \uIf{other does not contain method}{
                    $flag \longleftarrow False$
                }
                \Else{
                    \If{other.method.types does not fully contain method.types}{
                        $flag \longleftarrow False$
                    }
                }
            }

            \If{$flag = True$}{
                $relation[other].add$
            }
        }
    }
\end{algorithm}
%
\subsection{Polygons: Continued}
Let us illustrate the subtyping algorithm with our \emph{Polygons} \autoref{polygons}.
The algorithm generates the following subtyping relation, which also includes the subtyping relation for basic types defined in the first step.
\begin{flalign*}
    &Polygon = \left\{Polygon, Triangle, Square, Rhombus \right\} \qquad Triangle = \left\{Triangle\right\}\\
    &Square = \left\{Square\right\} \qquad  Rhombus  = \left\{Rhombus\right\} \qquad
    Diagonal = \left\{Diagonal\right\}
\end{flalign*}
Recall method \emph{diagonal} in class \emph{Diagonal}, we can see that the result of the set operation on its parameter type gives the following set of polygons:
\begin{flalign*}
    Polygon \textbf{ and not } Triangle = \left\{Polygon, Square, Rhombus \right\}
\end{flalign*}
If we pass to \emph{diagonal} an argument of type \emph{Triangle}, as it is not contained in the computed set of subtypes, we get a typechecking error at compile time \cite{UD20}.
